---
title: "R containers for Data"
author: "Kushal K Dey"
date: "July 28, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

There are several containers for data in R. Broadly they can be categorized into 4 gtoups 

- Vectors
- Matrices 
- Multi-dimensional arrays
- Lists
- Data Frame 

## Vectors

Create a vector (numeric or character in general).

```{r}
vec <- c(23, 14, 56, 76, 3, 2);
vec = c(23, 14, 56, 76, 3, 2); ## same as above
length(vec) ## gives you the size of the vector
class(vec) ## what kind of a vector is it?

vec <- c("Chicago", "Boston", "San Francisco", "New York")
class(vec)
length(vec)
```

Sequence vectors  (special cases)

```{r}
rep(2,length=10) ## repeat 2 10 times in a vector
seq(1,10, length.out=20) ## 20 uniformly spaced points between 1 and 10
seq(1,10,0.5) ## points at gaps of 0.01 from 1 to 10
1:15 ## sequential vector at gap 1 from 1 to 15 
```


### Basic operations on a vector

the other operations on a vector are as follows 

```{r}
vec <- c(23, 14, 56, 76, 3, 2);
mean(vec) ## mean of elements of the vector
median(vec) ## the middlemost value in the vector
sort(vec) ## sort all elements in the vector
order(vec) ## vec[order(vec)]= sort(vec) 
min(vec) ## minimum element of the vector
max(vec) ## maximum element of the vector
```

### More vector operations 

Other vector operations 

```{r}

vec <- c(23, 14, 56, 76, 3, 2);

## which element of vec is > 2
which(vec > 15)

## match another vector to the given vector
match(c(14, 3), vec)

## remove a set of elements from vec
setdiff(vec, c(14, 3, 23))

## intersection of two vectors
intersect(vec, c(14,3, 23))

## union of two vectors
union(vec, c(14,3, 23))

## elementwise transformation of vec

cat("log transformation of vec:", log(vec), "\n"); 
cat("square transformation of vec:", vec^2, "\n"); 
cat("square root transform of vec:", sqrt(vec), "\n")
cat("add 6 to vec", vec+6, "\n") 

```

### Two vectors comparison

We have a measure called correlation that tells us whether there is a strong positive or negative association between two vectors.Correlation is a measure between -1 and 1. It is -1 when there is a strong negative linear association, +1 when there is a strong positive linear association and 0 when there is no linear association.

```{r}
vec1 <- c(10, 20, 30, 40, 50);
vec2 <- c(2, 5, 10, 12, 20);
cor(vec1, vec2)

vec1 <- c(10, 20, 30, 40, 50);
vec2 <- c(20, 12, 10, 5, 2);
cor(vec1, vec2)

vec1 <- c(10, 20, 30, 40, 50);
vec2 <- c(4, 4.1, 3.2, 4.0, 3.8)
cor(vec1, vec2)
```

### sapply on vectors

The *sapply()* function which takes a vector as input and applied a user defined function on each element of the vector.

```{r}
vec <- c(23, 14, 56, 76, 3, 2);
sapply(vec, function(x) return(x^2 - x +2 )) 

## carries out this operation on each element of vec
```

## Matrices

Create a matrix (a 2D container for data)

One example

```{r}
mat <- matrix(0, nrow=2, ncol=2);## initialize with all 0
mat[1,1] <- 3
mat[1,2] <- 5
mat[2,1] <- mat[1,1] + mat[1,2];
```

Second example

Build a diagonal matrix from a vector

```{r}
diagmat <- diag(c(24,56,2))
```

### Basic operations on matrices

Matrix operations 

```{r}

## transpose of a matrix

t(mat)

## inverse of a mtrix

inv_mat <- solve(mat)
print(inv_mat)

## multiplication of two matrices

mat %*% inv_mat ## the dimensions must be proper

diag(mat) ## gives the diagonal entries of a matrix

mat[1,] ## access 1st row of a matrix
mat[,1] ## access 1st column of a matrix

vec1 <- c(12, 24, 56);
vec2 <- c(10, 3, 5);

mat <- cbind(vec1, vec2) ## bind vectors as columns
mat <- rbind(vec1, vec2) ## bind vectors as rows 

# solve(mat) ## will give ERROR (dimensions must be square)

```

### apply function on matrices

```{r}
## apply on the columns

apply(mat,2,mean)
apply(mat,2, function(x) min(x))

## apply on rows

apply(mat,1,sd)
apply(mat,1,function(x) max(x) - min(x))

## apply on each element

apply(mat, c(1,2), function(x) log(x))

```

## Mult-dimensional Arrays 

Beyond vectors which are 1D containers for data, and matrices which are 2D containers, we can have many-dimensional containers or tensors for data. 

```{r}
mult_mat <- array(0, c(2,3,2));
mult_mat[1,,] <- rbind(c(3,4), c(5,6), c(1,2));
mult_mat[2,,] <- rbind(c(5,2), c(10,0), c(1,6));
print(mult_mat)
dim(mult_mat)
```

The *apply()* function also works on these tensor objects.

```{r}
apply(mult_mat, c(2,3), mean)
apply(mult_mat, 1, function(x) sd(x))
```

## Lists

The vectors, matrices and arrays need data to be in structured form in an array format. For data that is not structured, we can use lists. Lists have the power of holding any types of objects, need not be vectors or matrices but even beyond (strings etc).

```{r}
lst <- list(0.5, 0.841, 0.977)
lst <- list(mid=0.5, right=0.841, top=0.977, bottom = 0.4)
lst[[1]]
lst$mid
lst$right
list_vec <- vector(mode="list");
list_vec[[1]] <- mult_mat;
list_vec[[2]] <- mat;
list_vec[[3]] <- vec2;
list_vec[[4]] <- 0

list_vec <- vector(mode="list");
list_vec[["first"]] <- mult_mat;
list_vec[["second"]] <- mat;
list_vec[["third"]] <- vec2;
list_vec[["fourth"]] <- 0

list_vec$second
list_vec$third
```

The *lapply()* function can be used to apply a function on each element of a list.

```{r}
lapply(list_vec, function(x) mean(x))
```

## Data Frme

A data frame is a matrix like representation of the data that stores all kinds of data, be it numeric, character, strings etc etc. We will see an example right here

```{r}
library(hflights)
data("hflights")
head(hflights)
```

