---
title: "Data Management in R"
author: "Kushal K Dey"
date: "July 27, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

### First steps

The first step in any data analysis is to read the data. 
\textcolor{red}{Caution}: *Note the size of the file (.csv or .txt or .xlsx) before reading the data*. If the data is less than 20-30 MB, use the very generic **read.csv**, **read.table** or *read_excel* function from the package *readxl*.

```{r echo=TRUE, eval=FALSE}
data <- read.table("PATH to FILE (.txt)")
data <- read.csv("PATH TO FILE (.csv)")
data <- readxl::read_excel("PATH TO FILE (.csv)")
```

If the size of the data is more than the above cut-off, use the *fread* function from the package **data.table**. 

```{r echo=TRUE, eval=FALSE}
data <- data.frame(data.table::fread("PATH TO FILE (.csv)"))
```

Now assuming you have loaded the data, lets look at how to handle a data.

We load a data on flight schedule from the **hflights** package. 

```{r}
suppressMessages(library(dplyr))
library(hflights)
data(hflights)
```

The data is now loaded and saved under *hflights*. The first thing you should do is check the dimension. 

```{r}
dim(hflights)
```

Caution: Do not do 

```{r echo=TRUE, eval=FALSE}
# hflights
```

if the data is big, there is a possibility that it may take a long time to print and you won't even be able to make much sense of the printed output.

Shortcuts around this problem. Look at a snapshot of the data.

```{r}
head(hflights) ## the top 6 rows 
tail(hflights) ## last 6 rows
glimpse(hflights) ## summary of the columns with types identified
summary(hflights)
```

### Local Data Frame

You can do a bunch of data handling and processing stuffs very elegantly in R, especially using the **dplyr** package. That is why it is always useful to convert a data frame to a local data frame using **dplyr**

```{r}
flights <- tbl_df(hflights)
```

If you need to print the data, do the following

```{r}
# convert to local data frame
flights <- tbl_df(hflights)

# printing only shows 10 rows and as many columns as can fit on your screen
flights

# you can specify that you want to see more rows
print(flights, n=20)

# convert to a normal data frame to see all of the columns
data.frame(head(flights))
```


###  Filtering Rows with matching criterion

```{r}
filter(flights, Month==1, DayofMonth==1)
```

If you want to save this filtered data, do

```{r}
filtered_data <- data.frame(filter(flights, Month==1, DayofMonth==1))
head(filtered_data) ## prints first few rows, match with previous
dim(filtered_data) ## dimension of the filtered data
```


```{r}
# use pipe for OR condition
filter(flights, UniqueCarrier=="AA" | UniqueCarrier=="UA")
```

It is same as 

```{r}
filter(flights, UniqueCarrier %in% c("AA", "UA"))
```

###  Selecting columns 

We can select columns simply by name 

```{r}
flights[, c("DepTime", "ArrTime", "FlightNum")] ## base - R function
```

or 

```{r}
select(flights, DepTime, ArrTime, FlightNum) ## dplyr function
```

```{r}
select(flights, Year:DayOfWeek)
select(flights, Year:DayofMonth, contains("Taxi"), contains("Delay"))
```

### Chaining or Pipelining 

```{r}
# nesting method to select UniqueCarrier and DepDelay columns and filter for delays over 60 minutes
filter(select(flights, UniqueCarrier, DepDelay), DepDelay > 60)
```

We can do the nesting in a sequential way 

```{r}
flights %>%
    select(UniqueCarrier, DepDelay) %>%
    filter(DepDelay > 60)
```

This improves the readability of the commands that should act sequential

```{r}
flights %>% 
  filter(DepDelay > 60) %>% 
  select(UniqueCarrier, DepDelay)  ## should be same as before
```

If we want to sort by *DepDelay*.

```{r}
## ascending order of DepDelay

flights %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(DepDelay)

## descending order of DepDelay

flights %>%
    select(UniqueCarrier, DepDelay) %>%
    arrange(desc(DepDelay))
```

### Add new columns 

We can add new columns processing from the current table using the `mutate` function.

```{r}
flights %>%
    select(Distance, AirTime) %>%
    mutate(Speed = Distance/AirTime*60)
```

A very useful technique is summarizing data by grouping a column.

`group_by` creates the group and `summarise` provides the aggregation function for the groups 

```{r}
flights %>%
    group_by(Dest) %>%
    summarise(avg_delay = mean(ArrDelay, na.rm=TRUE))
```

```{r}
# for each carrier, calculate the percentage of flights cancelled or diverted

flights %>%
    group_by(UniqueCarrier) %>%
    summarise_each(funs(mean), Cancelled, Diverted)

# for each carrier, calculate the minimum and maximum arrival and departure delays


flights %>%
    group_by(UniqueCarrier) %>%
    summarise_each(funs(min(., na.rm=TRUE), max(., na.rm=TRUE)), ArrDelay, DepDelay)

## same as before

flights %>%
    group_by(UniqueCarrier) %>%
    summarise_each(funs(min(., na.rm=TRUE), max(., na.rm=TRUE)), matches("Delay"))

```

### Sample from the data 

If we want to randomly select a few rows from the data, we can do it using the `sample_n` or `sample_frac` functions. 

```{r}
# randomly sample a fixed number of rows, without replacement
flights %>% sample_n(5)

# randomly sample a fraction of rows, with replacement
flights %>% sample_frac(0.25, replace=TRUE)

```

### More useful commands

We can use the `n()` and `n_distinct()` functionalities to determine the numebr of cases and the number of distinct cases in  agroup respectively. 

```{r}
# for each day of the year, count the total number of flights and sort in descending order

flights %>%
    group_by(Month, DayofMonth) %>%
    summarise(flight_count = n()) %>%
    arrange(desc(flight_count))

# rewrite more simply with the `tally` function

flights %>%
    group_by(Month, DayofMonth) %>%
    tally(sort = TRUE)

flights %>%
    group_by(Dest) %>%
    summarise(flight_count = n(), plane_count = n_distinct(TailNum))

```

Find rows corresponding to distinct values of a column using `distinct` function.

```{r}
flights %>%
  select(UniqueCarrier, FlightNum, TailNum) %>%
  distinct(FlightNum)
```

An example with multiple chaining commands 

```{r}
flights %>%
    group_by(UniqueCarrier) %>%
    select(Month, DayofMonth, DepDelay) %>%
    arrange(UniqueCarrier, desc(DepDelay))
```

