---
title: "Fundamentals of R"
author: "Kushal K Dey"
date: "July 28, 2016"
output: 
  html_document:
    css: floating-toc.css
    toc: true
---

##Installation Instructions

[http://statistics.uchicago.edu/~collins/Rinstall/](http://statistics.uchicago.edu/~collins/Rinstall/)

A generic way of installing and loading R packages 

```{r echo=TRUE, eval=FALSE}
install.packages("vegan") ## vegan is the name of the package
library(vegan)
```

## Directory path setting

Set the directory where you want to work. Preferably, same directory as the one where the script lies. 

```{r echo=TRUE, eval=FALSE}
setwd("PATH") ## set the path where you want to load or save files 
getwd()  ## gives back the directory you set
```

## Helper functions

If the user wants to look up the documentation for all functions of a package or a specific function of a package, he can type 

```{r echo=TRUE, eval=FALSE}
help(vegan) 
help(adonis, package="vegan") ## adonis is a function in vegan package
```

## Printing in R

Print statements 

```{r}
print("Hello World!!")
a <- 1
a
print(a)
print("a is equal to", a) ## Do not use
cat("a is equal to", a)
```

## Simplest mathematics

Basic functions

```{r}
a <- 2; b <- 4
c <- a+b  ## addition
c= a+b ## same as c <- a+b
d <- a - b ## subtraction 
e <- a*b ## multiplication
f <- a/b ## division
g <- a %% b ## modulo
exp(a); ## exponentiation
log(a); ## logarithm
a^2; ## square
sqrt(a) ## square root
```

## Conditional Statements  (if/else)

```{r}
counter = 0;

if (counter > 0){
  a=1
}else{
  a=0
}

a
```

## Looping (for and while loops)

```{r}
x <- 0;
for(i in 1:50){
  x <- x+i
  x
}

flag <- 1
while(flag < 4){
  print(flag)
  flag=flag+1
}

```

## Calling  a function 

```{r}
## Function call

## We want a function that computes a square of an object (may be numeric or a numeric vector)

prodvec <- function(x){
  x2 <- x*x
  return(x2)
}

prodvec(4)
prodvec(3:10)
```


## A random number generator

```{r}
runif(1,0,1) ## one random draw from 0 to 1
runif(100, 0, 1) ## 100 random draws from 0 to 1
runif(1,10,11) ## 1 random draw from 10 to 11
runif(10, -100, 100) ## 10 random draws from -100 to 100
```


